Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-04-01 talk/icooolps2009/paper.tex
give a quicky summary of the various phases of VMs before going into details
move the list after the assumptions
explain what is a trace

@@

When a hot loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations. During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.

>>>>

Typically, programs executed by a tracing VMs goes through various phases:

\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

The \emph{code generation} phase takes as input the trace generated during
\emph{tracing}.

@@

When a hot loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. \sout{When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations.
} \anto{During tracing, the interpreter records a history of all the
  operations it executes.}

\anto{
Such a history is called a \emph{trace}: it is a sequential list of
operations, togheter with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code: to ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example \texttt{if}s or indirect/virtual
calls.  When emitting the machine code, we turn every guard into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution in other
ways.  
}

During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that is being interpreted at the moment anyway.
